\documentclass[11pt]{article}
\usepackage{amsfonts,amsmath,amssymb,amsthm,graphicx}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage[caption=false]{subfig} 
\usepackage{color}
\usepackage{ifthen}

\setlength{\topmargin}{0 mm}
\setlength{\oddsidemargin}{5 mm}
\setlength{\evensidemargin}{5 mm}
\setlength{\textwidth}{150 mm}
\setlength{\textheight}{210 mm}

\include{macros}

\begin{document}

%% \title{Notes on Schroedinger's equation}

%% \author{N. Anders Petersson\thanks{Center for Applied
%%     Scientific Computing, Lawrence Livermore National Laboratory, L-561, PO Box 808, Livermore CA
%%     94551. }}

%% \date{\today}

%% \maketitle
\section{Hamiltonian structure}
Consider the scaled Schr\"odinger equation
\begin{equation}\label{eq_shrodinger}
\dot{\psi} = -i H(t) \psi,\quad \psi(0) = \psi_0,\quad H(t) = H_s + f(t)a + \bar{f}(t)a^\dag.
\end{equation}
Here, $\psi \in\mathbb{C}^N$ and $H_s = H_s^\dag$ is the system Hamiltonian matrix, which we assume to
be  Hermitian, real-valued, and independent of time. The lowering and raising matrices are denoted
$a$ and $a^\dag$, respectively. These matrices satisfy
\begin{equation}\label{eq_matrices}
%
a = \begin{bmatrix}
0 & 1 & & & &\\
 & 0 & \sqrt{2} & & &\\
&  & 0 & \sqrt{3} & &\\
& &  & 0 & \sqrt{4} & \\
& &  &  & \ddots & \ddots\\
\end{bmatrix},\quad
%
a^\dag = \begin{bmatrix}
0 &  & & &\\
1 & 0 & & &\\
&  \sqrt{2} & 0 &  &\\
& &  \sqrt{3} & 0 & \\
& &  & \ddots & \ddots
\end{bmatrix}
\end{equation}
The time-dependent control terms in \eqref{eq_shrodinger} can be written
\[
f(t)a + \bar{f}(t)a^\dag =
% (f_r(t) + i f_i(t)) a + (f_r(t) - i f_i(t)) a^\dag
f_r(t) (a + a^\dag) + i f_i(t) ( a- a^\dag),\quad f_r = \re f,\quad f_i=\im f,
\]
where $i=\sqrt{-1}$ is the imaginary unit. We conclude that the control terms are Hermitian and
$H(t) = H^\dag(t)$.

To utilize efficient numerical ODE solvers it is desirable to derive a real-valued equivalent of
\eqref{eq_shrodinger}. Let the functions $u(t)$ and $v(t) \in \mathbb{R} \to \mathbb{R}^N$ be
defined by
\[
\psi(t) = u(t) - iv(t)
\]
and decompose the total Hamiltonian matrix into $H(t) = K(t) + i S(t)$, where the real-valued
matrices $K$ and $S$ hold the symmeric and skew-symmetric parts of $H$: $K^\dag = K$ and $S^\dag =
-S$. We have,
\begin{align*}
H\psi &= (K+iS)(u - iv) = (Ku + Sv) + i(Su - Kv),\\
-iH\psi &= -i(Ku + Sv) + (Su - Kv).
\end{align*}
Therefore, a real-valued equivalent of the system \eqref{eq_shrodinger} is
\begin{equation}\label{eq_real-shrodinger}
  \begin{bmatrix} \dot{u}\\ \dot{v} \end{bmatrix} =
%
  \begin{bmatrix}
    S & -K \\ K & S
  \end{bmatrix}     
  %
  \begin{bmatrix} u\\ v \end{bmatrix} .
\end{equation}

We define the time-dependent Hamiltonian function to be
\begin{equation}\label{eq_hamiltonian}
\kappa(u,v,t) = u^T S(t) v + \frac{1}{2} u^T K(t) u + \frac{1}{2} v^T K(t) v.
\end{equation}
We have
\[
\frac{\p\kappa}{\p u_q} =
%
\frac{\p}{\p u_q} \sum_{j,k} \left( u_j S_{j,k} v_k + 0.5 u_j K_{j,k} u_k \right) =
%
\sum_{k} \left( S_{q,k} v_k + 0.5 K_{q,k} u_k \right) + 0.5 \sum_{j} u_j K_{j,q}.
\]
Because $K^T = K$,
\[
\nabla_u \kappa = S v + K u.
\]
In a similar way,
\[
\nabla_v \kappa = S^T u + K v = -S u + Kv.
\]
Therefore, \eqref{eq_real-shrodinger} has the Hamiltonian structure
\begin{equation}\label{eq_hamiltonian-form}
  \begin{bmatrix}
    \dot{u} \\   \dot{v}
  \end{bmatrix} =
  \begin{bmatrix}
    -\nabla_v \kappa(u,v,t)\\
    \nabla_u \kappa(u,v,t)
  \end{bmatrix}
\end{equation}
Because $v^TS^Tu = u^TSv$, the Hamiltonian function \eqref{eq_hamiltonian} can be written
\[
\kappa(u,v,t) = \frac{1}{2} \begin{bmatrix} u \\ v \end{bmatrix}^T
%
\begin{bmatrix}
  K(t) & S(t) \\ S^T(t) & K(t)
\end{bmatrix}
\begin{bmatrix} u \\ v \end{bmatrix}.
\]
We conclude that the Hamiltonian function is non-separable due to the skew-symmetric term
$S(t) =f_i(t) ( a- a^\dag)$ in the original Hamiltonian matrix.

\paragraph{Autonomous Hamiltonian form}
A standard approach for transforming a non-autonomous Hamiltonian system into autonomous form is as
follows. We first introduce the scalar dependent variables $e(t)$ and $\tau(t)$ and augment the
system \eqref{eq_hamiltonian-form} by the equations
\begin{align*}
  \dot{e} &= -\kappa_t(u,v,t),\\
  \dot{\tau} &= 1,\quad \tau(0)=0.
\end{align*}
The second equation has the solution $\tau(t) = t$. Thus we can write
$\kappa(u,v,t) = \kappa(u,v,\tau)$.
Next, we define the augmented dependent variables
\[
p = (u, e),\quad q=(v, \tau),
\]
and the extended Hamiltonian function
\[
\tilde{\kappa}(p,q) = \kappa(u,v,\tau) + e.
\]
We note that
\begin{align*}
  \nabla_{q} \tilde{\kappa} &= (\nabla_v \kappa, \kappa_\tau),\\
  \nabla_{p} \tilde{\kappa} &= (\nabla_u \kappa, 1).
\end{align*}
Therefore, the augmented variables satisfy
\[
\begin{bmatrix}
  \dot{p} \\
  \dot{q}
\end{bmatrix} =
%
\begin{bmatrix}
  (\dot{u}, \dot{e}) \\
  (\dot{v}, \dot{\tau})
\end{bmatrix} =
%
\begin{bmatrix}
  -\nabla_{q} \tilde{\kappa}(p, q)\\
  \nabla_{p} \tilde{\kappa}(p, q)
\end{bmatrix},
\]
which is an autonomous Hamiltonian system.  

\paragraph{$\rho$-reversibility UPDATE!}
Let $\rho$ be the linear invertible transformation
\[
\rho(p,q) = (p, -q).
\]
and denote $y=(p,q)$. The differential equation $\dot{y}=f(y)$ and the vector field $f(y)$ are
called $\rho$-reversible if
\[
\rho(f(y)) = -f(\rho(y)).
\]
Ignoring the time-dependence in the system \eqref{eq_hamiltonian-form}, we have
\[
f(y) =   \left( -\nabla_q \kappa(p,q), \nabla_p \kappa(p,q) \right) = ( S u - Kv, S v + K p ).
\]
Thus,
\[
f(\rho(y)) = \left( -\nabla_v \kappa(p,-q), \nabla_p \kappa(p,-q) \right) = ( S u + Kv, -S v + K u ),
\]
and
%% \[
%% \rho(f(y)) =  ( S p - Kv, -S v - K p ).
%% \]
%We conclude that the system \eqref{eq_hamiltonian-form} is {\em not} $\rho$-reversible due to the
%terms $Su$ and $Sv$.



\end{document}
